Union and Intersection Contracts Are Hard, Actually

### 2021-06-11 paper.tex
Pre-arxiv submission cleaning

\documentclass[sigplan,10pt,review,anonymous]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

\definecolor{OliveGreen}{RGB}{128,128,0}
%%%%%%%%%%%%%%%%% Editing marks %%%%%%%%%%%%%%%%%

  % TOGGLE ME to turn off all the commentary:
  \InputIfFileExists{no-editing-marks}{
    \def\noeditingmarks{}
  }

  \usepackage{xargs}
  \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
  % ^^ Need for pgfsyspdfmark apparently?
  \ifx\noeditingmarks\undefined
      \setlength{\marginparwidth}{1.2cm} % A size that matches the new PACMPL format
      \newcommand{\Red}[1]{{\color{red}{#1}}}
      \newcommand{\newaudit}[1]{{\color{blue}{#1}}}
      \newcommand{\note}[1]{{\color{blue}{\begin{itemize} \item {#1} \end{itemize}}}}
      \newenvironment{alt}{\color{red}}{}

      \newcommandx{\unsure}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{#2}}
      \newcommandx{\info}[2][1=]{\todo[linecolor=green,backgroundcolor=green!25,bordercolor=green,#1]{#2}}
      \newcommandx{\change}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{#2}}
      \newcommandx{\inconsistent}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
      \newcommandx{\critical}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=red,#1]{#2}}
      \newcommand{\improvement}[1]{\todo[linecolor=pink,backgroundcolor=pink!25,bordercolor=pink]{#1}}
      \newcommandx{\resolved}[2][1=]{\todo[linecolor=OliveGreen,backgroundcolor=OliveGreen!25,bordercolor=OliveGreen,#1]{#2}} % use this to mark a resolved question
  \else
  %    \newcommand{\Red}[1]{#1}
      \newcommand{\Red}[1]{{\color{red}{#1}}}
      \newcommand{\newaudit}[1]{#1}
      \newcommand{\note}[1]{}
      \newenvironment{alt}{}{}
  %    \renewcommand\todo[2]{}
      \newcommand{\unsure}[2][1=]{}
      \newcommand{\info}[2][1=]{}
      \newcommand{\change}[2]{}
      \newcommand{\inconsistent}[2]{}
      \newcommand{\critical}[2]{}
      \newcommand{\improvement}[1]{}
      \newcommand{\resolved}[2]{}
  \fi

%%%%%%%%%%%%%%%%% /Editing marks %%%%%%%%%%%%%%%%%

@@

% Copied from https://hal.inria.fr/hal-01140459/file/racket.inc.tex
\lstdefinelanguage{racket} {
  morekeywords=[1]{define, define-syntax, define-macro, lambda, define-stream, stream-lambda},
  morekeywords=[2]{begin, call-with-current-continuation, call/cc,
    call-with-input-file, call-with-output-file, case, cond,
    do, else, for-each, if,
    let*, let, let-syntax, letrec, letrec-syntax,
    define-context, define-controller, Integer, Boolean, get, when-required, when-provided,
    maybe_publish, require, submod, or/c, and/c, ->, \#\%module-begin,
    always_publish, with-syntax, define-struct/contract, syntax-case,
    define/contract,
    let-values, let*-values,
    module, provide,
    and, or, not, delay, force,
    \#`, \#',
    \#lang, implement, begin-for-syntax, rename-out,
    quasiquote, quote, unquote, unquote-splicing,
    map, fold, syntax, syntax-rules, eval, environment, query },
  morekeywords=[3]{import, export},
  alsoletter={',`,-,/,>,<,\#,\%},
  morecomment=[l]{;},
%  literate={lambda}{{\lambdaup}}1, % lambda -- look at https://tex.stackexchange.com/questions/119879/math-symbols-in-tt-font
  moredelim=**[is][\color{light-gray}]{<<@<<}{>>@>>},
  moredelim=**[is][\itshape\color{OliveGreen}]{<<;<<}{>>;>>},
  morecomment=[s]{\#|}{|\#},
  sensitive=true,
}

@@

\lstset{
  language={nickel},
  basicstyle=\small\ttfamily, % Global Code Style
  captionpos=b, % Position of the Caption (t for top, b for bottom)
  extendedchars=true, % Allows 256 instead of 128 ASCII characters
  tabsize=2, % number of spaces indented when discovering a tab
  columns=fixed, % make all characters equal width
  keepspaces=true, % does not ignore spaces to fit width, convert tabs to spaces
  showstringspaces=false, % lets spaces in strings appear as real spaces
  breaklines=true, % wrap lines if they don't fit
  frame=trbl, % draw a frame at the top, right, left and bottom of the listing
  frameround=tttt, % make the frame round at all four corners
  framesep=4pt, % quarter circle size of the round corners
  numbers=left, % show line numbers at the left
  numberstyle=\tiny\ttfamily, % style of the line numbers
  escapeinside={(*@}{@*)} % escape sequence to insert latex
  % commentstyle=\color{dark-grey}, % style of comments
  % keywordstyle=[1]\color{blue-portage}, % style of keywords
  % keywordstyle=[2]\color{orange-vivid-tangerine}, % style of keywords
  % keywordstyle=[3]\color{blue-portage}, % style of keywords
  % keywordstyle=[4]\color{green-pea}, % style of keywords
  % keywordstyle=[5]\color{pink-froly}, % style of keywords
  % stringstyle=\color{blue-marguerite}, % style of strings
}

\lstset{
  escapeinside={(*@}{@*)} % escape sequence to insert latex
}
\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
\newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}
\newcommand{\python}[1]{\lstinline[language=Python]{#1}}

\begin{abstract}
Union and intersection types are a staple of gradually typed language such
as TypeScript. While it's long been recognized that union and
intersection types are difficult to verify statically, it may appear
at first that the dynamic part of gradual typing is actually pretty
simple.

@@

\title{Union and intersection contracts are hard, actually}
\author{Teodoro Freund}
\affiliation{
  \institution{Universidad de Buenos Aires}
  \city{Buenos Aires}
  \country{Argentina}
}
\email{tfreund95@gmail.com}

@@

@@

On the other hand, unions are not a common feature of static type
systems, mostly because they are quite difficult to verify
statically. So unions are really only worth it in gradually typed
language where they formalise existing dynamically typed patterns. On
the other hand, surely, for dynamic tests, unions are really easy: it
is simply the Boolean disjunctions of two tests.

Unfortunately, as we document in this article, as soon as you extend
dynamic checks to \emph{contracts}~\cite{FindlerFelleisenHOContracts},
unions become actually pretty difficult, and threaten desirable
properties of your language.

\subsection{Configuration languages}

@@

In this article, we will use the Nickel language~\cite{NickelRepo} as
illustration and motivation. At its core, Nickel is the JSON data
model, augmented with abstraction mechanisms, and its conformed of:

\begin{itemize}
  \item dictionaries, written as:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
{field1 = value1, ..., fieldn = valuen}
\end{lstlisting}
  \item arrays:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
[x1, x2, ..., xn]
\end{lstlisting}
  \item functions:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
fun arg1 ... argn => body
\end{lstlisting}
  \item and let-definitions:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
let id = value in exp
\end{lstlisting}
\end{itemize}

% In this article, we will use the Nickel language~\cite{NickelRepo} as
% illustration and motivation. At its core, Nickel is the JSON data
% model, that is dictionaries (written \nickel{\{field1 = value1, ...,
%   fieldn = valuen\}}) and arrays (\nickel{[x1, ..., xn]}), augmented
% with functions (\nickel{fun arg1 ... argn => body}) and
% let-definitions (\nickel{let id = value in exp}).
\resolved{Arnaud: this
  paragraph looks kind of bad, maybe it'll be better to put each
  syntax entry in an enumeration, we will see if there is still room
  for this
  Teo: Tried to enumerate the constructs, left the original commented}

@@

\subsection{Contracts}

@@

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.\unsure{This point used to be made, do we want to
  reinsert it? However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code).}

@@

To be able to support such large dictionaries, Nickel's dictionaries
are \emph{lazy}, that is, the values are only evaluated when
explicitly required. For instance, when writing
\hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
evaluated.

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic checks
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be use to add schema validation to functions as well (in fact
functions were the original motivation for contracts).

There is no Boolean function which can check that a value has type
\nickel{Str -> Str}. Instead, a contract for \nickel{Str -> Str}
checks for each call of the function whether
\begin{enumerate}
\item the argument has type \nickel{Str}, otherwise the caller of the
  function is faulty
\item if so, that the returned value has type \nickel{Str}, otherwise
  the implementation of the function is faulty
\end{enumerate}

@@

\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
let catHosts = fun last =>
  let hosts = ["foo.com", "bar.org"] in
  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in

@@

    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let makeHost = fun server ext => [(*@{\ldots}@*)] in
  |                --------------------
  |                evaluated to this
  = ++, 2nd argument
\end{lstlisting}
    \caption{Error reporting without contract}
    \label{fig:contract-reporting-wo}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
let catHosts | Str -> Str = fun last =>
  [(*@{\ldots}@*)]
\end{lstlisting}
\vspace{10ex}
    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | --------------------------- (2) calling <func>
    \end{lstlisting}
    \caption{Error reporting with contract}
    \label{fig:contract-reporting-w}
  \end{subfigure}
  \caption{Contracts improve error messages}
\end{figure*}

Unfortunately, the delayed check of contract, while essential to
ensuring that schema validation doesn't affect performance (or, indeed,
is possible at all on functions), make union contracts (and their less
appreciated sibling, intersection contracts) quite problematic.

\subsection{Contributions}
Our contributions are as follows
\begin{itemize}
\item We describe the fundamental difficulties caused by presence of
  union and intersection contracts in a language, which are kept
  implicit in the literature (Section~\ref{sec:issues-sem})
\item We survey the various trade-offs which appear in implemented
  languages and in the literature to work around these difficulties
  (Section~\ref{sec:issues-literature})
\end{itemize}

\section{A typology of language features}
\label{sec:feat-lang}

@@

\subsection{User-defined contracts}
\label{sec:flat-contracts}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun p =>
  num.isInt p && 0 <= p && p <= 65535) in
80 | Port
\end{lstlisting}
\end{center}
\caption{A contract for valid ports}
\label{fig:port-contract}
\end{figure}

This type of contracts are present in many different languages,
for instance, the Eiffel programming language\cite{meyer1987eiffel}, the precursor
of the Design by Contract philosophy, makes it possible to assert
these kinds of expression as pre- and post-conditions on
functions and as invariants on classes\cite{EiffelDesignByContract}.

The Racket programming language also has a system to work with
contracts, powerful enough to define flat contracts, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts\cite{RacketContracts}.

\subsection{Referential transparency}
\label{sec:optimizations}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let subList = fun l1 l2 =>
  lists.any (fun x => x == (lists.head l1)) l2
  && subList (list.tail l1) l2
\end{lstlisting}
  \end{center}
\caption{Inlining}
\label{fig:optimizations-inlining-ex}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let elemAtOrLast = fun index list =>
  if index > lists.length list - 1 then
    lists.elemAt (lists.length list - 1) list
  else
    lists.elemAt index list
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let elemAtOrLast = fun index list =>
  let l = lists.length list - 1 in
  if index > l then
    lists.elemAt l list
  else
    lists.elemAt index list
\end{lstlisting}
  \end{center}
\caption{Common subexpression elimination}
\label{fig:optimizations-cse-ex}
\end{figure}

@@

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let f = fun x => g y (x + 1)
\end{lstlisting}
\begin{lstlisting}[language=nickel,title={Optimized program}]
let g' = g y in
let f' = fun x => g' (x + 1)
\end{lstlisting}
  \end{center}
\caption{Let-floating}
\label{fig:optimizations-let-floating-ex}
\end{figure}

@@

However, (\ref{eq:cbn-expansion}) does hold for \emph{pure} terms, that are
terms without side-effects. In a pure language, all these transformations are
valid.  In impure languages, the situation varies: in some case a large subset
of pure terms can be identified (in languages with effects tracking such as
PureScript) to be safely transformed.  Otherwise, the compiler must stay
conservative and only apply CSE to expressions it can prove are without
side-effects (arithmetic expressions, for example). \resolved{Cite a
different language, or remove citation.  (ADDED) tbh PureScript is still
statically-typed, compiled lang, but who knows: the "script" in the name and the
web orientation may still be more appealing to the audience of DLS}

Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
unions and intersections make (\ref{eq:cbn-expansion}) unsound \emph{even in a
pure setting}, preventing the kind optimization of
Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.

\section{Union \& intersection}
\label{sec:union-inter}

Let us now consider union and intersection contracts, before we
explain, in Section~\ref{sec:issues-sem} how they can compromise the
properties that we described in Section~\ref{sec:feat-lang}.

@@

\paragraph{In gradually typed systems}

The problem that these practical gradual type systems are trying to solve is to
capture, in static types, as many programming patterns as possible
from the underlying dynamically typed language (JavaScript for
TypeScript and Python for MyPy). One such pattern is
heterogeneous collections. For instance, in TypeScript, an array
which can contain both strings and numbers would have type
\typescript{Array<string|number>}.

@@

% Arnaud: I'm not deleting this quite yet because we may want to use
% these examples, I just didn't manage to make them fit in the flow of
% the narrative.
%
% Unions are not only useful to model the outcome of alternative
% branches of computations, but they can be used to
% formalize very common patterns of data types:
% \begin{itemize}
%     \item To represent nullable, or optional, values.
% \begin{lstlisting}[language=nickel]
% optionalValue | Str @| Null
% \end{lstlisting}

%     \item For plain enumerations of literals.
% \begin{lstlisting}[language=nickel]
% protocol | "Http" @| "Ftp" @| "Sftp"
% \end{lstlisting}

%     \item To support different representations of the same data.
% \begin{lstlisting}[language=nickel]
% person | {name : Str, age : Int}
%   @| {name : Str, dateOfBirth : Date}
% \end{lstlisting}

% \end{itemize}
% All of which are common in configurations.

\resolved{Arnaud: this paragraph is pretty good, on the other hand, there
is no value which lets me complete the example, so this example falls
a bit flat.
Teo: Improved the example, also added another one commented that is, for me,
more realistic, but I think the one uncommented is simpler.}

% \begin{figure}[h]
% \begin{lstlisting}[language=JavaScript]
% function addElem(values: Array<{ Nix: string }> | Array<{ Nickel: number }>) {
%     values.push({Nix: "Packages", Nickel: 32})
% }
% \end{lstlisting}
% \caption{Adding an element to a union of two arrays}
% \label{fig:addElem}
% \end{figure}

@@

Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
function \nickel{appendDate} append an element to \nickel{list} whose type is
only known to be in union of two lists, each using a different representation.
Both alternatives support falling back to a simple string for unparsed dates.
Faced with these two possibilities, \nickel{appendDate} can only append a
value which fits both types: this is precisely the intersection
\nickel{(Date @| Str) @& (DateWeek @| Str)}, that is, \nickel{Str}.

>>>>

\documentclass[sigplan,screen,10pt]{acmart}
%\usepackage{amssymb}
\usepackage{mathtools,amsmath,amsfonts,amsxtra,amsthm,mathrsfs,mathpartir}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{array}
\usepackage[toc,page]{appendix}
\usepackage{graphicx}
\usepackage{varioref}
\usepackage{verbatim,comment}
\usepackage{longtable}
\usepackage{tikz}
\usepackage{url}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{pbalance}

\definecolor{OliveGreen}{RGB}{128,128,0}

@@

% Copied from https://hal.inria.fr/hal-01140459/file/racket.inc.tex
\lstdefinelanguage{Racket} {
  morekeywords=[1]{define, define-syntax, define-macro, lambda, define-stream, stream-lambda},
  morekeywords=[2]{begin, call-with-current-continuation, call/cc,
    call-with-input-file, call-with-output-file, case, cond,
    do, else, for-each, if,
    let*, let, let-syntax, letrec, letrec-syntax,
    define-context, define-controller, Integer, Boolean, get, when-required, when-provided,
    maybe_publish, require, submod, or/c, and/c, ->, \#\%module-begin,
    always_publish, with-syntax, define-struct/contract, syntax-case,
    define/contract,
    let-values, let*-values,
    module, provide,
    and, or, not, delay, force,
    \#`, \#',
    \#lang, implement, begin-for-syntax, rename-out,
    quasiquote, quote, unquote, unquote-splicing,
    map, fold, syntax, syntax-rules, eval, environment, query },
  morekeywords=[3]{import, export},
  alsoletter={',`,-,/,>,<,\#,\%},
  morecomment=[l]{;},
%  literate={lambda}{{\lambdaup}}1, % lambda -- look at https://tex.stackexchange.com/questions/119879/math-symbols-in-tt-font
  moredelim=**[is][\color{light-gray}]{<<@<<}{>>@>>},
  moredelim=**[is][\itshape\color{OliveGreen}]{<<;<<}{>>;>>},
  morecomment=[s]{\#|}{|\#},
  sensitive=true,
}

@@

\lstset{
  language={Nickel},
  basicstyle=\small\ttfamily, % Global Code Style
  captionpos=b, % Position of the Caption (t for top, b for bottom)
  extendedchars=true, % Allows 256 instead of 128 ASCII characters
  tabsize=2, % number of spaces indented when discovering a tab
  columns=fixed, % make all characters equal width
  keepspaces=true, % does not ignore spaces to fit width, convert tabs to spaces
  showstringspaces=false, % lets spaces in strings appear as real spaces
  breaklines=true, % wrap lines if they don't fit
  % frame=trbl, % draw a frame at the top, right, left and bottom of the listing
    % frameround=tttt, % make the frame round at all four corners
  % framesep=4pt, % quarter circle size of the round corners
  numbers=none, % show line numbers at the left
  % numberstyle=\tiny\ttfamily, % style of the line numbers
  escapeinside={(*@}{@*)} % escape sequence to insert latex
  % commentstyle=\color{dark-grey}, % style of comments
  % keywordstyle=[1]\color{blue-portage}, % style of keywords
  % keywordstyle=[2]\color{orange-vivid-tangerine}, % style of keywords
  % keywordstyle=[3]\color{blue-portage}, % style of keywords
  % keywordstyle=[4]\color{green-pea}, % style of keywords
  % keywordstyle=[5]\color{pink-froly}, % style of keywords
  % stringstyle=\color{blue-marguerite}, % style of strings
}

\lstset{
  escapeinside={(*@}{@*)}, % escape sequence to insert latex
  numbers=none, % for the camera-ready version: numbers overlap with margins
  % Avoid overlap with margins
  % framexleftmargin = -2pt,
  % framexrightmargin = -2pt,
  % framextopmargin = -2pt,
  % framexbottommargin = -2pt
}

\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}
\newcommand{\haskell}[1]{\lstinline[language=Haskell]{#1}}
\newcommand{\typescript}[1]{\lstinline[language=JavaScript]{#1}}
\newcommand{\python}[1]{\lstinline[language=Python]{#1}}

\begin{abstract}
Union and intersection types are a staple of gradually typed languages such
as TypeScript. While it's long been recognized that union and
intersection types are difficult to verify statically, it may appear
at first that the dynamic part of gradual typing is actually pretty
simple.

@@

\title{Union and Intersection Contracts Are Hard, Actually}
\author{Teodoro Freund}
\affiliation{
  \institution{Universidad de Buenos Aires}
  \city{Buenos Aires}
  \country{Argentina}
}
\email{tfreund@dc.uba.ar}

@@

%%% The following is specific to DLS '21 and the paper
%%% 'Union and Intersection Contracts Are Hard, Actually'
%%% by Teodoro Freund, Yann Hamdaoui, and Arnaud Spiwack.
\setcopyright{acmlicensed}
\acmPrice{15.00}
\acmDOI{10.1145/3486602.3486767}
\acmYear{2021}
\copyrightyear{2021}
\acmSubmissionID{splashws21dlsmain-p31-p}
\acmISBN{978-1-4503-9105-4/21/10}
\acmConference[DLS '21]{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages}{October 19, 2021}{Chicago, IL, USA}
\acmBooktitle{Proceedings of the 17th ACM SIGPLAN International Symposium on Dynamic Languages (DLS '21), October 19, 2021, Chicago, IL, USA}

@@

On the other hand, unions are not a common feature of static type
systems, mostly because they are quite difficult to verify
statically. So unions are really only worth it in gradually typed
language where they formalise existing dynamically typed patterns.

Surely, for dynamic tests, unions ought to be really easy: they
are simply the Boolean disjunction of two dynamic tests, that fail
whenever one of those tests fail.
Unfortunately, as we document in this article, as soon as you extend
dynamic checks to \emph{contracts}~\cite{FindlerFelleisenHOContracts},
unions become actually pretty difficult, and threaten desirable
properties of your language.

\subsection{Configuration Languages}

@@

In this article, we will use the Nickel language~\cite{NickelRepo} as
illustration and motivation. At its core, Nickel is the JSON data
model, augmented with abstraction mechanisms, and it comprises:

\begin{itemize}
  \item dictionaries, written as
    \footnote{Note that, unlike JSON, Nickel assigns values in
    dictionaries using \nickel{=}, to keep \nickel{:} for type
annotations.}:
\begin{lstlisting}[frame=none,numbers=none,language=nickel]
{field1 = value1, ..., fieldn = valuen}
\end{lstlisting}
  \item arrays:
\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
[x1, x2, ..., xn]
\end{lstlisting}
  \item functions:
\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
fun arg1 ... argn => body
\end{lstlisting}
  \item and let-definitions:
\begin{lstlisting}[frame=none,numbers=none,language=Nickel]
let id = value in exp
\end{lstlisting}
\end{itemize}

@@

\subsection{Contracts}
\label{sub-sec:contracts}

@@

These are inherently dynamic questions, as they are all questions
about the evaluated configuration. To this effect, Nickel lets us
annotate any expression with a dynamic schema check: \nickel{exp |
  C}. There is also syntactic sugar to annotate definitions:
\nickel{let id | C = value in exp} stands for \nickel{let id = (value
  | C) in exp}.

@@

To be able to support such large dictionaries, Nickel's dictionaries
are \emph{lazy}, that is, the values are only evaluated when
explicitly required. For instance, when evaluating the expression
\hbox{\nickel{nixpkgs.hello},} only the \nickel{hello} package gets
evaluated, even if \nickel{nixpkgs} contained a
\nickel{world} package as well.

But let's consider now writing something like \nickel{nixpkgs |
  packages}, to guarantee that all the packages conform to the desired
schema. If this were a simple Boolean test, it would have to evaluate
all 50\,000 package to check their validity, hence breaking the
laziness of dictionaries. Do we have to choose between laziness and
schema validation? Fortunately, we don't! Enter
\emph{contracts}~\cite{FindlerFelleisenHOContracts}: dynamic checks
which can be partially delayed, yet errors can be reported
accurately. Contracts can respect laziness of dictionaries, and they
can be used to add schema validation to functions as well (in fact
functions were the original motivation for contracts).

There is no Boolean function which can check that a value has type
\nickel{Str -> Str}. Instead, a contract for \nickel{Str -> Str} \\
checks for each call of the function whether
\begin{enumerate}
\item the argument has type \nickel{Str}, otherwise the caller of the
  function is faulty
\item if so, that the returned value has type \nickel{Str}, otherwise
  the implementation of the function is faulty
\end{enumerate}

@@

\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=Nickel]
let catHosts = fun last =>
  let hosts = ["foo.com", "bar.org"] in
  lists.fold (fun val acc =>
    val ++ "," ++ acc) hosts last in

@@

    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let makeHost = fun server ext => [(*@{\ldots}@*)] in
  |                --------------------
  |                evaluated to this
  = ++, 2nd argument
\end{lstlisting}
    \caption{Error reporting without contract}
    \label{fig:contract-reporting-wo}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.48\linewidth}
    \begin{lstlisting}[language=nickel]
$let$ catHosts | Str -> Str $= fun last =>
  let hosts = ["foo.com", "bar.org"] in
  lists.fold (fun val acc =>
    val ++ "," ++ acc) hosts last in

catHosts (makeHost "google")$
\end{lstlisting}
    \begin{lstlisting}[frame=none,numbers=none, basicstyle=\footnotesize\ttfamily]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | --------------------------- (2) calling <func>
    \end{lstlisting}
    \caption{Error reporting with contract}
    \label{fig:contract-reporting-w}
  \end{subfigure}
  \caption{Contracts improve error messages}
\end{figure*}

As we shall show, the delayed check of contract, while essential to
ensuring that schema validation doesn't affect performance (or, indeed,
is possible at all on functions), make union contracts (and their less
appreciated sibling, intersection contracts) quite problematic.
While usual contracts require only one witness to show the invalidation
of a contract, the introduction of unions makes
the number of witnesses not bounded.

\subsection{Contributions}
Our contributions are as follows
\begin{itemize}
\item We describe the fundamental difficulties caused by pre-sence of
  union and intersection contracts in a language, which are kept
  implicit in the literature (Section~\ref{sec:issues-sem})
\item We survey the various trade-offs which appear in implemented
  languages and in the academic literature to work around these difficulties
  (Section~\ref{sec:issues-literature})
\end{itemize}

\section{A Typology of Language Features}
\label{sec:feat-lang}

@@

\subsection{User-Defined Contracts}
\label{sec:flat-contracts}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=Nickel]
let Port = contracts.fromPred (fun p =>
  num.isInt p && 0 <= p && p <= 65535) in
80 | Port
\end{lstlisting}
\end{center}
\caption{A contract for valid ports}
\label{fig:port-contract}
\end{figure}

This type of contracts are present in many different languages,
for instance, the Eiffel programming language~\cite{meyer1987eiffel}, the precursor
of the Design by Contract philosophy, makes it possible to assert
these kinds of expression as pre- and post-conditions on
functions and as invariants on classes~\cite{EiffelDesignByContract}.

The Racket programming language also has a system to work with
contracts, powerful enough to define user-defined contracts, and
to compose them with other kinds of dynamic checks,
like higher order contracts or a lightweight take on union
and intersection contracts~\cite{RacketContracts}.

\subsection{Referential Transparency}
\label{sec:optimizations}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=Nickel,title={Source program}]
let elem = fun elt =>
  lists.any (fun x => x == elt) in

let subList = fun l1 l2 =>
  elem (lists.head l1) l2
  && subList (list.tail l1) l2
\end{lstlisting}
\begin{lstlisting}[language=Nickel,title={Optimized program}]
let subList = fun l1 l2 =>
  lists.any (fun x => x == (lists.head l1))
            l2
  && subList (list.tail l1) l2
\end{lstlisting}
  \end{center}
\caption{Inlining}
\label{fig:optimizations-inlining-ex}
\end{figure}

@@

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=Nickel,title={Source program}]
let elemAtOrLast = fun index list =>
  if index > lists.length list - 1 then
    lists.elemAt (lists.length list - 1)
                 list
  else
    lists.elemAt index list
\end{lstlisting}
\begin{lstlisting}[language=Nickel,title={Optimized program}]
let elemAtOrLast = fun index list =>
  let l = lists.length list - 1 in
  if index > l then
    lists.elemAt l list
  else
    lists.elemAt index list
\end{lstlisting}
  \end{center}
\caption{Common subexpression elimination}
\label{fig:optimizations-cse-ex}
\end{figure}

@@

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=Nickel,title={Source program}]
let f = fun x => g y (x + 1)
\end{lstlisting}
\begin{lstlisting}[language=Nickel,title={Optimized program}]
let g' = g y in
let f' = fun x => g' (x + 1)
\end{lstlisting}
  \end{center}
\caption{Let-floating}
\label{fig:optimizations-let-floating-ex}
\end{figure}

@@

However, (\ref{eq:cbn-expansion}) does hold for \emph{pure} terms, that are
terms without side-effects. In a pure language, all these transformations are
valid.  In impure languages, the situation varies: in some case a large subset
of pure terms can be identified (in languages with effects tracking such as
PureScript) to be safely transformed.  Otherwise, the compiler must stay
conservative and only apply CSE to expressions it can prove are without
side-effects (arithmetic expressions, for example).

Strikingly, we will see in Section~\ref{sec:issues-sem} that the introduction of
union and intersection contracts \emph{breaks referential transparency} and make
(\ref{eq:cbn-expansion}) \emph{unsound}\footnote{(\ref{eq:cbn-expansion}) holds
  in Nickel despite non-termination and contract-checking errors
  because of lazy evaluation. In strict languages (\ref{eq:cbn-expansion}) needs to be restricted, nevertheless
  union and intersection contracts make it worse.}, preventing the kind of optimization of
Figure~\ref{fig:optimizations-let-floating-ex} to fire in general.

\section{Union \& Intersection}
\label{sec:union-inter}

Let us now consider union and intersection contracts, before we explain in
Section~\ref{sec:issues-sem} how they can compromise the properties that we
described in Section~\ref{sec:feat-lang}.

@@

\paragraph{In Gradually Typed Systems}

The problem that these practical gradual type systems are trying to solve is to
capture, in static types, as many programming patterns as possible
from the underlying dynamically typed language (Java\-Script for
TypeScript and Python for MyPy). One such pattern is
heterogeneous collections. For instance, in TypeScript, an array
which can contain both strings and numbers would have type
\typescript{Array<string|number>}.

@@

@@

Figure~\ref{fig:addElem} gives a concrete example of this phenomenon. The
function \nickel{appendDate} appends an element to \nickel{list}, whose type is
only known to be the union of two lists, each using a different representation.
Because the return type is the same as the input type, \nickel{appendDate} must
preserve this representation (\nickel{Date} and \nickel{DateWeek}
cannot be mixed in a same list).  Both alternatives support falling back to
a simple string for unparsed dates.  Faced with these two possibilities,
\nickel{appendDate} can only append a value which fits both types: this is
precisely the intersection \nickel{(Date @| Str) @& (DateWeek @| Str)}, that
is, \nickel{Str}.
